Details matter: Similar-looking host galaxies can produce different merging compact binary populations

### 2015-05-28 paper/paper.tex
paper and bib: editing to add in theory of BH mass and spin distributions

\newcommand\mc{{\cal M}_c{}}
\newcommand\editremark[1]{{\color{red}#1}}
\newcommand\msun{M_\odot}
\newcommand\unit[1]{\text{#1}}
\newcommand\abbrvPSgrb{PSG}
\newcommand\abbrvPSellipticals{PSE}
\newcommand\abbrvKBLowZa{BD2010}
\newcommand\ForInternalReference[1]{}
\newcommand\modelDefault{A}
\newcommand\modelNiino{A'}
\newcommand\modelKW{D}
\newcommand\modelMannucci{D'}
\newcommand\modelKB{B}
\newcommand\modelPanter{C}
\newcommand\cqg{CQG}
%\newcommand\aap{A\&A}
%\newcommand\apss{APSS}
%\newcommand\aaps{AAPS}
%% \newcommand\apjs{ApJ S}
%% \newcommand\aj{AJ}
%% \newcommand\apjl{ApJL}
%% \newcommand\mnras{MNRAS}
%% \newcommand\pasp{PASP}
%% \newcommand\pasj{PASJ}
%% \newcommand\araa{ARAA}
%% \newcommand\physrep{Phys. Rep.}

@@

  The simulation h258 (Exciting Galaxy), on the other hand, has a much
  more active merger history.  At $z = 1$ there is a 1:1 merger event;
  a gas disk rapidly reforms following the collision (see
  \citet{Governato09}), resulting in a massive disk galaxy at $z = 0$
  which looks remarkably similar to the Milky Way and to the other
  simulation, h277.  Prior to the $z = 1$ merger, each of the two
  progenitor galaxies actually experiences its own additional major
  merger events around $z = 2.5$ ({\bf check!}).  The combination of
  the series of major mergers, plus a number of minor interactions and
  flybys, gives a stark contrast to the relatively quiescent history
  of h277.  At $z = 0$, h258 has a virial mass of $M_{vir} = 7.74
  \times 10^{11} \msun$, stellar mass $M_* = 4.46 \times 10^{10}$
  \msun, and circular velocity $v_{circ} = \editremark{xx}$ km s$^{-1}$.

@@

% POINT: Trend versus mass and opening angle.  [Be careful: disruption time will depend on opening angle]
To estimate whether a specific compact binary will be tidally disrupted prior to merger, we employ the
following procecure to identify posterior samples consistent with tidal disruption of a neutron star.  First, the
smaller mass must lie within the range of masses allowed by our fiducial NS equation of state (\textbf{MPA1 from Ben}), here
between $0.5 M_\odot$ and $2.5 M_\odot$.  Second, the disruption process must leave behind a remnant disk with nonzero
mass, as estimated using Foucart's expression \cite{2012PhRvD..86l4007F} (his Eqs. (6,12-13)):
\begin{align}
\frac{M_{disk}}{M_{NS}} &=
 0.288 (3 m_{bh}/m_{ns})^{1/3}[1-2 {\cal C}] 
\nonumber \\
&
 -0.148 (m_{bh}/m_{ns})  {\cal C} R_{isco}(a)/m_{bh} 
\label{eq:Mdisk}
\end{align}
where $q=m_{bh}/m_{NS}$ is the binary mass ratio; $a=S_{bh}/m_{bh}^2$ is a dimensionless measure of the black hole spin;
and  where ${\cal C}(m_{ns})=m_{ns}/R_{ns}$ is a mass- and equation-of-state dependent  measure of the neutron star
compactness.  In Foucart's expression, shown above, $R_{isco}(a)$ is the radius of the  innermost stable equitorial circular orbit  of a test particle
about a Kerr black hole  \cite{1972ApJ...178..347B}. % Brandon:  [their Eq. 2.21].   
When the black hole spin is not aligned with the orbit, we use the aligned component of the black hole spin magnitude $a\rightarrow a \cos \theta_{LS}$ in this
expression.   Fixing the neutron star mass $m_{ns}=1.4 M_\odot$ and hence compactness to ${\cal C}=1.66$, this expression implies a two-dimensional surface in
the three-dimensional parameter space $m_{bh},a,\theta_{LS}$ seperates parameters corresponding to tidal disruption to
parameters corresponding to prompt coalescence.
%
After coalescence, the remnant disk generates a tightly beamed electromagnetic counterpart.  For simplicity, we assume
beaming angle is independent of the disruption process or progenitor parameters, so a constant beaming correction factor $f_b$ enters into all
subsequent calculations and can hence be omitted.  
%
Figure \ref{fig:TidalDisrupt} shows contours of this expression for different values of $\theta_{LS}$.

\begin{figure}
\caption{\textbf{Disruption fraction}: Approximate tidal disrpution fraction}
\end{figure}

% POINT: Distribution of tilt angles
Based on Monte Carlo simulations of supernova kicks \editremark{citations:Eggleton,Kalogera,IhmXXX}, we approximate the tilt angle distribution $p(\theta)$ of bound,
coalescing BH-NS binaries by the positive half of a normal distribution
\[
p(\theta)\simeq e^{-\theta^2/2\sigma_\theta^2} \frac{2}{\sqrt{2\pi \sigma_\theta^2}}
\]
where $\sigma_{\theta}\simeq 1$ is a characteristic angular width that varies relatively weakly with the black hole mass
and supernova kick velocity, for velocities consistent with pulsar recoil velocities \editremark{citations}.  This simple distribution is consistent with some detailed population synthesis simulations by
\texttt{StarTrack}. \editremark{validate with more modern startrack}  
For simplicity, in this work we adopt $\sigma_{\theta}=1$.   
%  POINT Average over spin distributions
Using this approximate distribution of misalignment angles, Figure  assumed black hole mass and spin we calculate the
fraction of coalescing binaries consistent with

% POINT: Average over distributions

The relative probability of a given tilt angle

- first aproximations: tilt angle independent of merger time

@@

------

\editremark{calibrate mass distribution to synthetic universe/Dominik - but also allow reversed trend}
DISCUSSION: could we explain the redshift distribution of short GRBs this way?

@@

** mass-cut-dependent event rate -- will be more sensitive (e.g., if there
  is a hard cutoff on some masses, as having no EM counterpart)
\end{verbatim}

@@

\appendix
\section{Tidal disruption probability}

>>>>

\newcommand\mc{{\cal M}_c{}}
\newcommand\editremark[1]{{\color{red}#1}}
\newcommand\msun{M_\odot}
\newcommand\unit[1]{\text{#1}}
\newcommand\abbrvPSgrb{PSG}
\newcommand\abbrvPSellipticals{PSE}
\newcommand\abbrvKBLowZa{BD2010}
\newcommand\ForInternalReference[1]{}
\newcommand\modelDefault{A}
\newcommand\modelNiino{A'}
\newcommand\modelKW{D}
\newcommand\modelMannucci{D'}
\newcommand\modelKB{B}
\newcommand\modelPanter{C}
\newcommand\cqg{CQG}
\newcommand\E[1]{\left<#1\right>}
%\newcommand\aap{A\&A}
%\newcommand\apss{APSS}
%\newcommand\aaps{AAPS}
%% \newcommand\apjs{ApJ S}
%% \newcommand\aj{AJ}
%% \newcommand\apjl{ApJL}
%% \newcommand\mnras{MNRAS}
%% \newcommand\pasp{PASP}
%% \newcommand\pasj{PASJ}
%% \newcommand\araa{ARAA}
%% \newcommand\physrep{Phys. Rep.}

@@

  The simulation h258 (Exciting Galaxy), on the other hand, has a much
  more active merger history.  At $z = 1$ there is a 1:1 merger event;
  a gas disk rapidly reforms following the collision (see
  \citet{Governato09}), resulting in a massive disk galaxy at $z = 0$
  which looks remarkably similar to the Milky Way and to the other
  simulation, h277.  Prior to the $z = 1$ merger, each of the two
  progenitor galaxies actually experiences its own additional major
  merger events around $z = 2.5$ ({\bf check!}).  The combination of
  the series of major mergers, plus a number of minor interactions and
  flybys, gives a stark contrast to the relatively quiescent history
  of h277.  At $z = 0$, h258 has a virial mass of $M_{vir} = 7.74
  \times 10^{11} \msun$, stellar mass $M_* = 4.46 \times 10^{10} \msun$, and circular velocity $v_{circ} = \editremark{xx}$ km s$^{-1}$.

@@

% POINT: Trend versus mass and opening angle.  [Be careful: disruption time will depend on opening angle]
To estimate whether a specific compact binary will be tidally disrupted prior to merger, we employ the
following procecure to identify posterior samples consistent with tidal disruption of a neutron star.  First, the
smaller mass must lie within the range of masses allowed by our fiducial NS equation of state (\textbf{MPA1 from Ben}), here
between $0.5 M_\odot$ and $2.5 M_\odot$.  Second, the disruption process must leave behind a remnant disk with nonzero
mass, as estimated using the expression in \cite{2012PhRvD..86l4007F} (his Eqs. (6,12-13)):
\begin{align}
\frac{M_{disk}}{M_{NS}} &=
 0.288 (3 m_{bh}/m_{ns})^{1/3}[1-2 {\cal C}] 
\nonumber \\
&
 -0.148 (m_{bh}/m_{ns})  {\cal C} R_{isco}(a)/m_{bh} 
\label{eq:Mdisk}
\end{align}
where $q=m_{bh}/m_{NS}$ is the binary mass ratio; $a=S_{bh}/m_{bh}^2$ is a dimensionless measure of the black hole spin;
and  where ${\cal C}(m_{ns})=m_{ns}/R_{ns}$ is a mass- and equation-of-state dependent  measure of the neutron star
compactness.  In Foucart's expression, shown above, $R_{isco}(a)$ is the radius of the  innermost stable equitorial circular orbit  of a test particle
about a Kerr black hole  \citep{1972ApJ...178..347B}. % Brandon:  [their Eq. 2.21].   
When the black hole spin is not aligned with the orbit, we use the aligned component of the black hole spin magnitude $a\rightarrow a \cos \theta_{LS}$ in this
expression.   Fixing the neutron star mass $m_{ns}=1.4 M_\odot$ and hence compactness to ${\cal C}=1.66$, this expression implies a two-dimensional surface in
the three-dimensional parameter space $m_{bh},a,\theta_{LS}$ seperates parameters corresponding to tidal disruption to
parameters corresponding to prompt coalescence.
%
After coalescence, the remnant disk generates a tightly beamed electromagnetic counterpart.  For simplicity, we assume
beaming angle is independent of the disruption process or progenitor parameters, so a constant beaming correction factor $f_b$ enters into all
subsequent calculations and can hence be omitted.  
%
Figure \ref{fig:TidalDisrupt} shows contours of this expression for different values of $\theta_{LS}$.

\begin{figure}
\includegraphics{Figures/fig-mma-PrecomputeDisrupt}
\caption{\label{fig:TidalDisrupt:Probability}\textbf{Disruption fraction}: Contour plot of the fraction
  $P_{dis}(m_{bh},\chi_{bh})$ of BH-NS binaries with BH mass and spin $m_{bh},\chi_{bh}$ that are expected to tidally
  disrupt using the assumptions in the text.  Contours are shown at $0.1, 1/3, 0.5, 0.9$.}
\end{figure}

% POINT: Distribution of tilt angles
Based on Monte Carlo simulations of supernova kicks as described in the Appendix, we approximate the
cumulative tilt angle distribution $P(<\theta)$ of bound,
coalescing BH-NS binaries by the positive half of a normal distribution
\begin{eqnarray}
P(<\theta)\simeq \int_0^\theta e^{-\theta^2/2\sigma_\theta^2} \frac{2}{\sqrt{2\pi \sigma_\theta^2}} d\theta
\end{eqnarray}
where $\sigma_{\theta}\simeq 1$ is a characteristic angular width that varies relatively weakly with the black hole mass,
 supernova kick velocity, and compact binary formation channel.  This simple distribution is consistent with some detailed population synthesis simulations by
\texttt{StarTrack}.
For simplicity, in this work we adopt $\sigma_{\theta}=1$.   
%  POINT Average over spin distributions
Using this approximate distribution of misalignment angles, Figure \ref{fig:TidalDisrupt:Probability} estimates the fraction
$P_{dis}(m_{bh},\chi_{bh})$ of coalescing binaries with a given black hole mass and spin that should tidally disrupt.  
For simplicity assuming detection weighting does not strongly favor specific relative orientations $\theta$, this
distribution determines the relative likelihood that a gravitational wave event with a given BH mass and spin produces
is consistent with an electromagnetic counterpart.

% POINT: Average over distributions
The relative event rate of electromagnetic counterparts to BH-NS disruptions therefore depends on the intrinsic
distribution $p_d(m_{bh},\chi_{bh})$ of black hole mass and spin among GW-detected BH-NS mergers and the probability
that an event is detected via electromagnetic emission and GW: 
\begin{eqnarray}
p_{em}(m_{bh},\chi_{bh}) = \frac{P_{dis}(m_{bh},\chi_{bh}) p_d(m_{bh},\chi_{bh})}{\int dm_{bh} d\chi_{bh}P_{dis}(m_{bh},\chi_{bh}) p_d(m_{bh},\chi_{bh})}
\end{eqnarray}
where the constant beaming factor cancels.  
%
Because black holes in BH-NS binaries are currently expected to accrete relatively little mass, the spin distribution at merger
should be close to their birth spin distribution.   This distribution is entirely unknown but likely significant, given
tidally-induced spinup prior to merger\editremark{Ott}; we will assume the spin distribution to be uniform between $0.7$
and $1$, independent of formation channel or birth conditions: \editremark{change: make this one Z dependent}
\begin{eqnarray}
p(\chi_{bh}) = \frac{1}{0.3} \theta(\chi_{bh}-0.7)\theta(1-\chi_{bh})
\end{eqnarray}
As discussed in \cite{popsyn-LowMetallicityImpact2-StarTrackRevised-2012},  the black hole mass distribution depends
sensitively on the birth environment and coalescing binary formation channels; see their Tables 7 and 9 and their Figures 6,9, 11, and 13,
reconstructing the BH mass from the reported chirp masses (red curves).  
Omitting formation channels that predict implausibly high compact binary formation rates (their model A), the
distribution of $m_{bh}$ in BH-NS binaries extends over a wide range.  As a first approximation, the BH mass
distribution in BH-NS binaries does not depend on composition, favoring BH-NS binaries with relatively low BH masses
$m_{bh}\lesssim 5-9 M_\odot$ independent of birth composition.  
% \cite{2012ApJ...749...91F}  : physics of SN explosions used in these mechanisms. Rapid engine needed for mass gap, but
% is there one? 
% mc[m1_, m2_] := (m1 m2)^(3/5)/(m1 + m2)^(1/5)
% Solve[mc[m1, 1.4] == 2.5]
For simplicity, we will adopt a uniform mass distribution between $5$ and $9 M_\odot$:
\begin{eqnarray}
p(m_{bh}) = \frac{1}{4 M_\odot} \theta(m_{bh}-4 M_\odot)\theta(9 M_\odot-m_{bh})
\end{eqnarray}

@@

@@

** mass-cut-dependent event rate -- will be more sensitive (e.g., if there
  is a hard cutoff on some masses, as having no EM counterpart)

* Aside: BBH distribution the most sensitive, but not relevant for EM followup
\end{verbatim}

@@

\appendix
\section{Binary evolution and spin-orbit misalignment}

\editremark{ROS to do} -- find python/C code (Davide?) for simple MC; mine old resuts from startrack

%% POINT: Other reference
% 2014arXiv1401.3032W  - fallback details

The progenitors of both components of a BH-NS binary undergo supernova, whose associated asymmetric mass loss and
supernova kicks significantly perturb the orbit  \citep{2013PhRvD..87j4028G,2002MNRAS.329..897H,2000ApJ...541..319K}.   Assuming progeniror stellar spins are
aligned with the orbital plane (e.g., due to strong tides), only asymmetric supernova kicks can change
the orbital plane and introduce misalignment between the BH spin and the orbit.   The distribution of spin-orbit
misalignments expected in any specific binary depends strongly on weakly-constrained binary evolution physics, the
treatment of which is beyond the scope of this article.  Previous detailed calculations have reported on spin-orbit
misalignment \citep{2010ApJ...719L..79F,2008ApJ...682..474B,2004PhRvD..69j2002G}.  That said, relatively simple order-of-magnitude calculations
can explain the typical small misalignment angles expected among merging BH-NS binaries.
% POINT: Small angle tilt forma (follows from simple considerations of angular momentum)
Specifically, in the limit of a small supernova kick $\vec{v}_K$, the orbital plane  change by an angle $\theta \simeq \vec{v}_K \cdot
\hat{z}/v_{orb}$ where $\hat{z}$ is the direction perpendicular to the (circular) orbital plane.  The black hole forms first and cannot
significantly change its spin magnitude or direction through accretion; therefore the two supernovae kicks produced by
forming the black hole and then neutron star add (in quadrature) to produce a characteristic RMS misalignment error
$\E{\theta^2}$.
%
For simplicity, we will adopt the model used in the \texttt{StarTrack} binary evolution code \citep{StarTrack2}: binaries
prior to each supernovae can be approximated as in circular orbits;  supernova kicks on neutron stars produce a Maxwellian
distribution of velocities with 1d velocity dispersion $\sigma_v=265$ \citep{HobbsKicks};  supernova kicks on black holes
$v_{bh}$ are suppressed by fallback, with velocities  $\vec{v}_{K,BH} = (1-f_b)v_{NS}$ suppressed by fallback accretion
onto the black hole; and the dominant formation channels for BH-NS binaries involve common
envelope accretion after the first supernova forms a black hole.   Combining these expressions under these assumptions, we expect the net misalignment  $\theta$ due to both supernova
explosions will have zero mean and variance
\begin{eqnarray}
\label{eq:MisalignmentOrderOfMagnitude}
\E{\theta^2} = \frac{\E{v^2}}{3}\left[ \frac{(1-f_b)^2\epsilon_1}{v_{orb,1}^2}+\frac{\epsilon_2}{v_{orb,2}^2} \right]
\end{eqnarray}
where $v_{orb,1}$ and $v_{orb,2}$ are the orbital velocities at the first and second supernova and where
$\epsilon_1,\epsilon_2$ are factors between 0 and 1 which account for the fact that the average is taken over binaries which \emph{survive} and are not
disrupted by the kick.  Because the second supernova  occurs in a tight orbit (post common
envelope),  the orbital velocity $v_{orb,2}$ at the supernova producing the neutron
star will be  large:  if $M_2$ and $r_2$ are the total mass and orbital separation, $v_{orb,2}=\sqrt{GM_{2}/r}\simeq
440\unit{km/s} \sqrt{M/M_\odot/(r_2/R_\odot)}$, 
% Convert[Sqrt[  GravitationalConstant (mbh SolarMass)/(r SolarRadius)], (Kilo Meter/    Second)]
comparable to or
significantly in excess of the typical supernova kick velocity for massive black holes orbiting the stripped progenitors of
neutron stars at seperations of a few $R_\odot$.  Conversely, at birth the black hole will often capture a significant
amount of the stellar envelope (``fallback'' accretion or prompt collapse) with $f_b\simeq 1$.  Therefore, even though
the initial orbit is wide, orbital  velocity is very small, and hence the orbit can be easily strongly misaligned by
supernova kicks,  the first kick when acting on a massive black hole can produce little to  modest amounts orbital
misalignment.

% POINT: Second term of order unity
For exceptional models of binary evolution, the population of coalescing BH-NS binaries will depend extremely critically on the
a finely-tuned second supernova.  In this case, only a small fraction of surviving BH-NS binaries merge; the BH-NS merger rate will be extremely low and  depend extremely sensitively on
the magnitude of supernova kicks; and the misalignment produced by the second supernova can frequently be very large.  More typically, however, a significant fraction of surviving BH-NS binaries merge
\citep{2008ApJ...675..566O,popsyn-LowMetallicityImpact2-StarTrackRevised-2012,popsyn-LowMetallicityImpact2b-StarTrackRevised-2013,popsyn-LowMetallicityImpact2c-StarTrackRevised-2014}; in these cases, the  second supernova's kick does not require fine-tuning to guarantee
coalescence.   In these most common scenarios, the second term in Eq. (\ref{eq:MisalignmentOrderOfMagnitude}) is of
order unity.

% POINT:
As a result, the process of surviving two supernova kicks generally produces a wide  distribution of misalignment
angles concentrated near $\theta\simeq 0$.  This wide distribution is expected to be relatively robust.  Extremely wide
distributions are disfavored kinematically and by the requirement that sufficiently many binaries survive to coalesce.
Conversely, the width of the distribution is bounded below by the contribution from the second supernova kick.\footnote{For simplicity and brevity, we neglect the
  possibility of different modes of supernova explosions and hence kicks, notably via electron capture supernovae \citep{2012ARNPS..62..407J,2004ApJ...612.1044P,2010MNRAS.406..656K,2015ApJ...801...32A}.  That said, small supernova kicks in NS
  birth would further minimize any spin-orbit misaligment and hence further favor the strong dependence of tidal disruption probability on the black hole mass and spin
  distribution described in the text.} 
As a first approximation,  the cumulative distribution of misalignment angles should agree with direct calculations \cite{2010ApJ...719L..79F,2008ApJ...682..474B,2004PhRvD..69j2002G}.

% POINT: 
%% To calculate the necessary inputs to this expression, we adopt one particular family of supernova mechanism:  Fryer et al 
%% \cite{2012ApJ...749...91F}, the model adopted in \texttt{StarTrack} and applied in detailed calculations of binary
%% evolution
%% \cite{popsyn-LowMetallicityImpact2-StarTrackRevised-2012,popsyn-LowMetallicityImpact2b-StarTrackRevised-2013,popsyn-LowMetallicityImpact2c-StarTrackRevised-2014}.
%% Because of significant residual uncertainty in the supernova mechanism, our results are proof-of-concept calculations
%% intended to illustrate plausible effects of composition on remnant masses, spins, and hence misalignments.  Other models
%% for supernovae produce dramatically different initial-final mass relations \cite{2012ApJ...757...69U} and are
%% anticipated to produce equally different different amounts of spin-orbit
%% misalignment in BH-NS binaries.  
%% %
%% Observations of isolated pulsar
%% recoils \cite{1997MNRAS.291..569H,Ar} and binary pulsars \cite{psr-kicks-WongWillemsKalogera2010}
%% strongly constrain the action of random supernova kicks on neutron stars.
